Improvement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-08 rigid_bodies.tex
Revision of the NVT integration section

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega$ $=$ $\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

The factorization which played a central role in Ref.~\citenum{Silveira_2017} is ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, where ${\mt A}(\vt q)$ is the rotation matrix that connects the body-fixed and space-fixed frames.

The Hamiltonian conserved by Eq.~\eqref{eq:ODE system for NVE} is
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

@@

\subsection{NVT Dynamics with a Single Thermostat Chain: Notation and Measure-Preserving Integration}
\label{sec:nvt}
In this work, we couple a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} to both the translational and rotational degrees of freedom of the rigid-body system.
To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momenta $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$.
The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the target temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.

@@

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
\end{align}
\end{subequations}
With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.

If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are integrals of motion, while the system's center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if one makes $L = 6N$.
In the particular situation in which one sets $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of a simulation, one must make $L = 6N - 3$ \cite{Martyna_1994}.

Altough it is not possible to devise symplectic integrators for Eq.~\eqref{eq:ODE system for NVT}, one can rigurously obtain measure-preserving numerical solvers \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure $e^{-w(\vt x)}d\vt x$ for Eq.~\ref{eq:ODE system for NVT} is given by $w = -7N \eta_1 - \sum_{j=2}^M \eta_j$.
It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics so, in this case, the measure-preserving condition for every individual operator could be relaxed.

By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
The operator can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}
In the scheme above, the first propagator to act is $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, which promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ act sequentially, with $j$ varying from $M-1$ down to $1$.
With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as \cite{Martyna_1996}
\begin{align*}
&p_{\eta_j} = p_{\eta_j}^\ast + \Big( G_j^\ast - \alpha_{j+1}^\ast p_{\eta_j}^\ast \Big) \phi\left(\frac{\alpha_{j+1}^\ast \timestep}{4n}\right) \frac{\timestep}{4n} \; \text{and} \\
&\eta_{j+1} = \eta_{j+1}^\ast + \alpha_{j+1}^\ast \frac{\timestep}{4n}.
\end{align*}

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$.
Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
This numerical scheme is simply a measure-preserving version of the NVT integrator introduced by \citeauthor{Martyna_1996} \cite{Martyna_1996}.

For having a low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort, making it feasible to increase the time step $\timestep$ without degrading the accuracy of the NHC part.
In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
Here we choose to take a simpler path.
Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied deviations, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when computing ensemble averages.

>>>>

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

The factorization which played a central role in Ref.~\citenum{Silveira_2017} is ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, where ${\mt A}(\vt q)$ is the rotation matrix that connects the body-fixed and space-fixed frames. Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

@@

\subsection{NVT Dynamics with a Single Thermostat Chain: Notation and Measure-Preserving Integration}
\label{sec:nvt}

In this work, we couple a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} to both the translational and rotational degrees of freedom of the rigid-body system.
To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j = 1, \cdots, M$.
The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an \textit{extended energy} given by \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the target temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.

@@

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
\end{align}
\end{subequations}

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as a driven variables \cite{Tuckerman_2001}.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of \citeauthor{Tuckerman_2001} \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.

Eq.~\eqref{eq:ODE system for NVT} defines an invariant measure in the extended phase-space \cite{Tuckerman_1999}, and it is possible to devise numerical solvers that preserve such measure \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
We have done it by adapting the integrator introduced by \citeauthor{Martyna_1996} \cite{Martyna_1996} in light of a criterion explained by \citeauthor{Ezra_2006} \cite{Ezra_2006}.
It is worth noting, however, that such adaptation required us to alter the integration of the $\eta$ coordinates only.
Since these are driven variables, with no influence on the dynamics of the system, we can conclude that the practical importance of such alteration is small.

By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of Sec.~\ref{sec:nve} and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
This one can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

In the scheme above, the first propagator to act is $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, which promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ act sequentially, with $j$ varying from $M-1$ down to $1$.
With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as \cite{Martyna_1996}
\begin{align*}
&p_{\eta_j} = p_{\eta_j}^\ast + \Big( G_j^\ast - \alpha_{j+1}^\ast p_{\eta_j}^\ast \Big) \phi\left(\frac{\alpha_{j+1}^\ast \timestep}{4n}\right) \frac{\timestep}{4n} \; \text{and} \\
&\eta_{j+1} = \eta_{j+1}^\ast + \alpha_{j+1}^\ast \frac{\timestep}{4n}.
\end{align*}

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$.
Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.

Due to its low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort.
This makes it feasible to increase the time step $\timestep$ without ruining the accuracy of the NHC part.
In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
\citeauthor{Omelyan_2008}'s processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
Here we choose to take a simpler path, though.
Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble average are computed.
